Color- and piece-blind chess

### 2019-03-09 trunk/chess/papers/blind.tex
Wrote about using unblind to play

Everything was trained on a single GPU, though a fairly decent one in
2018, the EVGA GeForce GTX 1080 FTW (really), using OpenCL. Biases
were initialized to $0$ and weights with a Gaussian of mean $0$ and
standard deviation $0.025$. In the first phase, there were 64 examples
per round, and after vacuuming, 2048. The round learning rate
$\alpha_r$ started at $0.1$ and descended linearly to $0.002$ over
500,000 rounds; and the learning rate when updating weights (for each
example) $\alpha$ is $\alpha_r / \verb+examples_per_round+$. In no way
am I recommending these parameters. Fiddling with the parameters to
make it do its black magic or alternately carom off to a sea of NaNs
or zeroes is for sure the worst part of neural networks. In fact, I
initially started with the classical sigmoid transfer function, but
``upgraded'' to the ``leaky rectified linear'' function
%
$$((potential < 0.0f) ? potential * 0.01f : potential)$$
%
after getting fed up with sigmoid weights caroming off (see
``vanishing gradient problem'' and/or ``exploding gradient problem'').
The final model was trained over 339,885 rounds on 223 million
examples. It did not appear to show signs of improving for
several days before I terminated it.

@@

The unblinding component can be evaluated on its own, by running
it on examples sampled independently of the training set. Over
50,000 examples, these were the results:

@@

A few things are distinctly disappointing about its performance. Even
outside of ``likely'' positions, it usually predicts that pieces on
black's side of the board are black, and vice versa
(Figure~\ref{fig:allon}). This makes sense, but suggests serious
limitation on using the prediction to play chess. Less forgivably, it
sometimes predicts more than one king per side (or zero), which is
always wrong. Actually, an early version had this problem in spades,
usually frequently predicting two or three kings. Upon debugging, I
had simply made the noob mistake of printing both King and Knight with
the letter ``K.'' Ha! It often predicts the ``wrong'' number of
bishops (etc.), or places them on the same color. This is technically
possible through promotion, but usually a bad guess, since promotion
is relatively rare, and moreso promotion to a piece other than a
queen. An approach that might handle this better (but may have
different downsides) would be instead to predict the ``fates'' of the
32 initial pieces. The fate of a pawn can include promotion, but no
other pieces can. This would {\em require} that the model only predict
a single location for each king, among other things. However, this
would require a much larger output layer (32 pieces can move to 64
squares, plus promotion) and it is not always clear how to interpret
its value into a position (for example, if two pieces are predicted
strongly to be on the same square).

@@

>>>>

Everything was trained on a single GPU, though a fairly decent one in
2018, the EVGA GeForce GTX 1080 FTW (really), using OpenCL. Biases
were initialized to $0$ and weights with a Gaussian of mean $0$ and
standard deviation $0.025$. In the first phase, there were 64 examples
per round, and after vacuuming, 2048. The round learning rate
$\alpha_r$ started at $0.1$ and descended linearly to $0.002$ over
500,000 rounds; and the learning rate when updating weights (for each
example) $\alpha$ is $\alpha_r / \verb+examples_per_round+$. In no way
am I recommending these parameters. Fiddling with the parameters to
make it do its black magic or alternately carom off to a sea of NaNs
or zeroes is for sure the worst part of neural networks. In fact, I
initially started with the classical sigmoid transfer function, but
``upgraded'' to the ``leaky rectified linear'' function
%
$$(p < 0) \,?\, p \times 0.01 : p$$
%
after getting fed up with sigmoid weights caroming off (see
``vanishing gradient problem'' and/or ``exploding gradient problem'').
The final model was trained over 339,885 rounds on 223 million
examples. It did not appear to show signs of improving for
several days before I terminated it.

@@

The unblinding component can be evaluated on its own, by running it on
examples sampled independently of the training set. The model outputs
a score for each possible contents of each square; we simply
discretize to the highest-scoring one (same too for the predicted
castling state and turn). Over 50,000 examples, these were the
results:

@@

A few things are distinctly disappointing about its performance. Even
outside of ``likely'' positions, it usually predicts that pieces on
black's side of the board are black, and vice versa
(Figure~\ref{fig:allon}). This makes sense, but suggests serious
limitation on using the prediction to play chess. Less forgivably, it
sometimes predicts more than one king per side (or zero), which is
always wrong. Actually, an early version had this problem in spades,
usually frequently predicting two or three kings. Upon debugging, I
had simply made the noob mistake of printing both King and Knight with
the letter ``K.'' Ha! It often predicts the ``wrong'' number of
bishops (etc.), or places them on the same color. This is technically
possible through promotion, but usually a bad guess, since promotion
is relatively rare, and moreso promotion to a piece other than a
queen. An approach that might handle this better (but may have
different downsides) would be instead to predict the ``fates'' of the
32 initial pieces~\cite{survival}. The fate of a pawn includes what if
any piece it has promoted to, but no other pieces can. This would {\em
  require} that the model only predict a single location for each
king, among other things. However, this would require a much larger
output layer (32 pieces can move to 64 squares, plus promotion) and it
is not always clear how to interpret its value into a position (for
example, if two pieces are predicted strongly to be on the same
square).

@@

\section{Playing blind}

Once we have predicated a board state, we can play with it. The
simplest way to do this is to use a strong chess engine to pick a move
for the position. Here, I use Stockfish with a node budget of 1
million nodes, which takes about 1 second of CPU time per move on my
computer; henceforth {\sf stockfish1m}.

There are some complications:
\begin{itemize}
\item Frequently, the unblinded board will not match the true
  position, and Stockfish will choose a move that is illegal. So, as discussed
  before, we actually return a prioritzed list of moves. For this first
  experiment, we just return the move Stockfish recommends, followed by all
  other moves ordered randomly.
\item Stockfish is a very strong engine, and in my opinion the code is generally
  good, but it is very sensitive to bad FEN (the notation used to give it a
  position to evaluate) strings. Given a bad string, like one that says castling
  is possible when the king isn't in its home square, often crashes the engine.
  So we need to make sure to actually pass valid positions. I accomplish this
  by making the following modifications to the predicted board:
  \begin{itemize}
  \item If a castling bit is set, but castling is not possible due to the position
    of the king or rook, I clear the bit.
  \item Set the side-to-move to be the correct actual value. This uses
    the unblinded state, so is superficially cheating. But note that
    if we get the side wrong, then Stockfish's move will always be
    illegal: Moves are specified as a source and destination
    square,\footnote{Plus promotion piece. Castling is represented as
      a two-square move of the king.} and so the source square of
    Stockfish's move would always be a piece of the wrong piece's
    color. So this is equivalent to (but twice as efficient as)
    running stockfish twice, one for each side, and prioritizing the
    predicted side's move first.
  \end{itemize}
  This won't fix all positions, for example, if the white and black king are
  adjacent to one another in mutual check. If an irreparable problem is detected,
  then I just return a uniformly random move list.
\end{itemize}

It is easy to beat this chess engine, by tricking it as in
Figure~\ref{fig:trick}, although this involves unnatural moves, so it
may only apply if you know how it works.
% XXX example game?
Measuring how well it plays in an absolute sense is a subject of some
interest, so I wrote a separate paper about that~\cite{eloworld}. This
algorithm, called \verb+blind_yolo+, had an Elo World score of XXX.

\subsection{We three kings}

When evaluating the first version I found that it was producing a
fairly high number of illegal positions, which was causing us to fall
back on making random moves, which is mostly boring. The second version
reduces the rate of illegal positions due to too many or too few kings.

 - we get scores with each piece
 etc.

This version, called \verb+blind_kings+, performs consistently better
than \verb+blind_yolo+ (XXX give actual record). It had an Elo World
score of XXX.

\cite{kingme}

\section{Conclusion}

- Other structures for the network
- Stochastic approach
- How do the solutions compare to other chess algorithms that
  could be applied in the blind? Random works, also lexicographic,
  any others? Most are inspecting the board state, obviously.
- Could estimate the upper bound for accuracy by blinding every
  game in the database, and doing a ``lookup''-based version of
  the eval.
